Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-09 paper.tex
Nixed prompt tags and just went with shift/reset.

\section{What to expect}
This paper lays out a common framework to talk about pushdown analysis
for higher-order languages. In particular, we extract the ``essence''
of the existing techniques into a quality of the machine semantics
that specify a language's meaning, much in the same way as
~\citet{dvanhorn:VanHorn2012Systematic} (a method we will refer to as
AAM: abstracting abstract machines). Once the machine semantics is in
this form, simple pointwise abstraction leads to the summarization
algorithms that we see in the literature. In effect, we give a
\emph{concrete} semantics to pushdown analysis.

Summarization algorithms need not be restricted to languages with
well-bracketed calls and returns. We can adopt the technique for
higher precision in the common case but still handle difficult cases
such as first-class control. This was shown for the
\rackett{call-with-current-continuation} (a.k.a. \rackett{call/cc})
operator in ~\citet{ianjohnson:Vardoulakis2011Pushdown}. This
impressive work illuminated the fact that we can harness the enhanced
technology of pushdown analyses in non-pushdown models of
computation. Doing this sacrifices call/return matching in the general
case, but in practice the precision is much better than the
alternative regular\footnote{Regular as in regular language.} model
that, say, AAM would provide. A downside of the work is that it was an
algorithmic change to the already complicated CFA2 - there was no
recipe for how to do this for one's favorite control operator. Instead
of deriving a ``pushdown'' analysis for a language with
\rackett{call/cc}, we will show a new analysis for a language with the
all the control operators in
\citet{ianjohnson:Flatt:2007:ADC:1291151.1291178} (call it the PLT
machine) in order to demonstrate the applicability of this viewpoint
even in the context of complex control operators.

The remaining sections of the paper are
\begin{itemize}
\item{section \ref{sec:pdcfa}: we derive a cousin of PDCFA}
\item{section \ref{sec:cfa2}: we make additions to the previous semantics to get a direct-style CFA2 without first-class control}
\item{section \ref{sec:plt}: we give a novel analysis of the PLT machine, a core calculus of Racket's control operations.}
\end{itemize}

\section{Deriving PDCFA}
\label{sec:pdcfa}
PDCFA is a simpler analysis than CFA2, and thus enjoys a polynomial
complexity in the monovariant case. Its final ${\mathcal O}(n^6)$ form
unfortunately is not a complete abstraction of an unbounded stack
model. In this section, we show how to derive a ${\mathcal O}(n^9)$
complete abstraction\footnote{Interestingly, the available
  \emph{implementations} of PDCFA incorporate the fix and are
  themselves ${\mathcal O}(n^9)$.}. These bounds are absolute worst
case that would never ever happen in practice. The complete
abstraction restricts value flow to fewer spurious paths, often
leading to a faster analysis than a typical 0CFA.

The magic of the method is in keeping a table of (local) continuations
for each function $\times$ store pair. This closely mirrors the
technique of store-allocating continuations used in AAM. Instead of
deferring to the table (store) for the remainder of a continuation for
each frame, we only have indirections at function call
boundaries. Note that since AAM is about finitizing the state space,
this step itself fits within the AAM method, since continuations
within functions truncated at call boundaries are bounded by the
nesting depth of those functions, thereby making the continuation
space finite. In the monovariant case, the number of continuations is
still linear in the size of the program.

The tail of a continuation in the context of a function will contain
the stackless context in which the function was called, in order to
link up with the proper callsite(s). The context includes the function
(or unique label of the function), the environment, and the store. If
doing a flow analysis that takes advantage of the pushdown model, the
context would also have a lattice element.

Once we finitize the address space of the store, this computes a
characteristic finite state machine that has only reachable states of
the pushdown system the unbounded stack model embodies.

A ``summary edge'' that is in the literature, in terms of pushdown
systems, is an edge from the source of a push edge to the target of
the matching pop edge. ``Matching'' here means there is a path through
machine edges that don't change the stack, or through summary
edges. There is an analogy to something more operational: summary
edges embody memoization. Instead of following an entire path through
a call to return a value, we simply jump from the call to the return
(the result of the call). We are very conservative with how much of
the context is necessary to memoize, so we use the entire heap. This
conservativeness isn't useful for skipping any work in computing the
flow analysis (so much so that we don't even bother consulting the
memo table in the semantics), but it is useful for interpreting the
results to talk about the behavior of the body of a function without
detours to other functions. We will return to this analogy and what it
means in the presence of first-class control in section \ref{sec:plt}.

\begin{figure}
  \begin{align*}
    \mexpr \in \Expr &= \svar{\mvar} \mid \sapp{\mexpr}{\mexpr} \mid \slam{\mvar}{\mexpr} \\
    \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
    \mval \in \Value &::= (\slam{\mvar}{\mexpr},\ \menv) \\
    \mkont \in \Kont &::= \kmt \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
    \menv \in \Env &= \Var \to \Addr \\
    \mstore \in \Store &= \Addr \to \wp(\Value) \\
    \Var &\text{ an infinite set} \\
    \Addr &\text{ an infinite set}
  \end{align*}
  \caption{The CESK semantic spaces}
\label{fig:cesk-spaces}
\end{figure}

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\slam{\mvar}{\mexpr}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

\begin{figure}
  \begin{align*}
    \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \times \KTab \times \Memo \\
    \mkont \in \Kont &::= \kmt \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \mid \krt{(\mexpr,\menv),\mstore}\\
    \KTab &= (\Expr \times \Env)\times \Store \to \wp(\Kont) \\
    \Memo &= (\Expr \times \Env)\times \Store \to \wp(\Value)
  \end{align*}
  \caption{Extended semantic spaces}
\label{fig:cesk-ext-spaces}
\end{figure}

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\slam{\mvar}{\mexpr}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

To turn this semantics into PDCFA's algorithm for constructing a Dyck
state graph, we apply a widening operator ${\mathcal F}$ to make $\mstore$, $\mkont$,
and $\mmemo$ shared amongst all states. Further techniques for a
performant implementation can be found in
\citet{ianjohnson:oaam:2013}.

\begin{align*}
  \mastate \in \sa{State} &= \Expr \times \Env \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S', (\mastate',\mstore') \notin S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo) &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. Vardoulakis and
Shivers had a clear goal of harnessing the extra information a
pushdown model provides to produce a high-precision analysis that
works well in practice. This resulted in more than just the
call/return matching of the previous section, which is why we are
showing the two separately. There are two extra features of the
semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved nondeterminism}
\end{enumerate}
The first of these is an addition to the stackless context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have proved to
decrease running time in most cases.

The second of these is to ameliorate a problem they call ``fake
rebinding.'' That is, since bindings in the abstract represent several
values, we don't want to reference a variable \texttt{x} in two
different places and have it resolve to two different values. In AAM,
a variable reference non-deterministically steps to all possible
values associated with that variable. Here we want to say that once
\texttt{x} is considered to stand for value $v$, then all subsequent
references of \texttt{x} should be $v$. If they aren't, it looks as if
\texttt{x} was rebound; it hasn't, and thus it is a ``fake
rebinding.'' CFA2 does not step to all values on variable reference,
but instead carries all its values around in superposition until they
need to be observed at, say, a function call. We give a simplified
semantics that is more along the AAM style, but CFA2's approach can
easily be recovered from it.

We show only the significantly modified rules of the semantics in
figure \ref{fig:frame-semantics}. The other rules simply carry along
the extra $\mframe$ component untouched.
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar[\mlab]{\mvar}, \menv), \mstore, \mframe, \mkont}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$ if $(\mframe', \mval) \in \lookup(\mstore,\mframe,\menv,\mvar,\mlab)$
    \\
% % application
%     $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mframe, \mkont}$
%     &
%     $\tpl{(\mexpri0, \menv), \mstore, \mframe, \kar{\mexpri1,\menv,\mkont}}$
%     \\
% % argument evaluation
%     $\tpl{\mval, \mstore, \mframe, \kar{\mexpr,\menv,\mkont}}$
%     &
%     $\tpl{(\mexpr, \menv), \mstore, \mframe, \kfn{\mval, \mkont}}$
%     \\
% function call
    $\tpl{\mval,\mstore,\mframe,\kfn{\slam{\mvar}{\mexpr}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore', \mframe', \mkont}$
    \\ where & $(\mstore',\mframe') = \bind(\mstore,\mframe,\maddr,\mvar,\mval)$
  \end{tabular}
  \caption{The CES$\xi$K machine}
  \label{fig:frame-semantics}
\end{figure}

@@

\section{Analysis of the PLT machine}
\label{sec:plt}

@@

Composable and non-composable continuations both have their uses, as
it has been shown the two are not co-expressive when it comes to space
complexity ~\citet{ianjohnson:dybvig:control:2007}. There is still
merit in delimiting the stack in either case, as is argued by
\citet{ianjohnson:Flatt:2007:ADC:1291151.1291178}. Their paper gives a
semantics for the coexistence of composable and non-composable
continuations in the presence of first-class prompts,
\rackett{dynamic-wind} (protects entry/exit of a continuation with
pre- and post-processors), and continuation marks. We will show how to
apply pushdown analysis techniques in the presence of all this
complexity.

The way we have been splitting continuations at function calls has
similarities to the meta-continuation approach to modeling delimited
control ~\citep{ianjohnson:Biernacki2006274}. We could view each
function call as inserting a prompt, and returns as aborting to the
nearest prompt. This view contends with tail calls, but we can
identify tail calls easily - any call with a prompt as its
continuation is a tail call. Tail calls are important in language
implementations for space complexity reasons
\citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these
concerns are less important. The repeated popping of prompts inserted
by what otherwise were tail-calls is synonymous with CFA2's
``transitive summaries.''

We first give a CESK semantics of the PLT machine with the allocation
procedure as a parameter in order to first discuss the complexities
any computable abstract semantics will have. We then present the key
ideas to making it a table-based semantics, and how we can abstract it
to a computable approximation.

\subsection{PLT's CESK semantics}

>>>>

\section{Introduction}

Programs in higher-order languages heavily use function calls and
method dispatch as part of their control flow. Until recently, flow
analyses could not handle return flow precisely
\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:earl2010pdcfa}, which
leads to several spurious paths (and thus false positives) due to
their pervasiveness. These works, called CFA2 and PDCFA respectively,
use pushdown automata as their approximation's target model of
computation. They are hence called ``pushdown analyses.''\footnote{We
  will refer to the classic finite model analyses as ``regular
  analyses'' after regular languages.} The precision and performance
gains of these analyses are so beneficial that we believe new analyses
going forward should be using pushdown techniques. CFA2 and PDCFA have
difficult details to easily apply to an off-the-shelf semantics ---
especially if they feature non-local control transfer that breaks the
pushdown model.

...

\section{What to expect}

CFA2 uses a technique called ``summarization'' from
\citet[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with the
$\epsilon$-closure graph that PDCFA constructs. Summarization
algorithms need not be restricted to languages with well-bracketed
calls and returns. We can adopt the technique for higher precision in
the common case but still handle difficult cases such as first-class
control. This was shown for the
\rackett{call-with-current-continuation} (a.k.a. \rackett{call/cc})
operator in ~\citet{ianjohnson:Vardoulakis2011Pushdown}. This
impressive work illuminated the fact that we can harness the enhanced
technology of pushdown analyses in non-pushdown models of
computation. Doing this sacrifices call/return matching in the general
case, but in practice the precision is much better than the
alternative regular model that, say, AAM would provide.

This paper lays out a common framework to talk about pushdown analysis
for higher-order languages. In particular, we extract the ``essence''
of the existing techniques into a quality of the machine semantics
that specify a language's meaning, much in the same way as
~\citet{dvanhorn:VanHorn2012Systematic} (a method we will refer to as
AAM: abstracting abstract machines). Once the machine semantics is in
this form, simple pointwise abstraction leads to the summarization
algorithms that we see in the literature. In effect, we give a
\emph{concrete} semantics to pushdown analysis.

A downside of the work providing \rackett{call/cc} is that it was an
algorithmic change to the already complicated CFA2 --- there was no
recipe for how to do this for one's favorite control operator. This
paper seeks to do just that with an operational view of what
summarization is, in essence. In other words, we give a concrete
semantics to the tricks that the analyses in the literature use in the
abstract. We also give intuitive analogies to well-established
ideas/techniques so that you too can write a pushdown analysis for
your language. In order to demonstrate the applicability of this
viewpoint, we show a new analysis for a language with composable
control. All of the semantics modeled in this paper are implemented in
full detail in PLT Redex and available at

\begin{center}
  \url{http://github.com/ianj/concrete-summaries}
\end{center}

The remaining sections of the paper are
\begin{itemize}
\item{section \ref{sec:pdcfa}: we derive a cousin of PDCFA}
\item{section \ref{sec:cfa2}: we make additions to the previous semantics to get a direct-style CFA2 without first-class control}
\item{section \ref{sec:sr}: we give a novel analysis of delimited and composable first-class control.}
\end{itemize}

\section{Deriving PDCFA}
\label{sec:pdcfa}

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision, so we start by deriving PDCFA from an
operational semantics for the untyped lambda calculus (figure
\ref{fig:base-semantics}). The semantic spaces for the machine follow.

\begin{align*}
  \mexpr \in \Expr &= \svar{\mvar} \mid \sapp{\mexpr}{\mexpr} \mid \slam{\mvar}{\mexpr} \\
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \sclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
\end{align*}

The $\Addr$ space is what controls the precision of the model. For a
concrete semantics, we require the allocation metafunction $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin
\dom(\mstore)$), but any choice of address is sound.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\sclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

Since PDCFA is a simpler analysis than CFA2, it enjoys a polynomial
complexity in the monovariant case. Its final ${\mathcal O}(n^6)$ form
is a naive ``step all states'' widening, which yields poor
results. The common widening uses a worklist to only step states that
need stepping --- the fact that the store is shared amongst all states
makes this a semantic difference, and not just an implementation
difference. Doing this bumps the complexity up to ${\mathcal
  O}(n^9)$\footnote{The available \emph{implementations} of PDCFA
  indeed do this and are themselves ${\mathcal O}(n^9)$.}. These
bounds are (rough) absolute worst case that is highly unlikely, if
ever, happen in practice. The worklist abstraction restricts value
flow to fewer spurious paths, often leading to a faster analysis than
a typical 0CFA.

\paragraph{Tracking return points:} the magic of the method is in
keeping a table of (local) continuations for each function and store
pair. This closely mirrors the technique of store-allocating
continuations used in AAM. Instead of deferring to the table (store)
for the remainder of a continuation for each frame, we only have
indirections at function call boundaries. Note that since AAM is about
finitizing the state space, this step itself fits within the AAM
method, since continuations within functions truncated at call
boundaries are bounded by the nesting depth of those functions,
thereby making the continuation space finite. In the monovariant case,
the number of continuations is still linear in the size of the
program.

The tail of a continuation in the context of a function will contain
the stackless context in which the function was called, in order to
link up with the proper call-site(s). The context includes the
function (or unique label of the function), the environment, and the
store. Notice that this choice can be viewed as a special allocation
strategy for continuations in the AAM viewpoint; the store is the
important ingredient to maintaining enough precision to keep a
pushdown abstraction. We separate the concepts in this paper for the
exposition. The semantic spaces for the machine are modified thusly:

\begin{align*}
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \times \KTab \times \Memo \\
  \mkont \in \Kont &::= \kmt
                     \mid \kar{\mexpr,\menv,\mkont} 
                     \mid \kfn{\mval,\mkont}
                     \mid \krt{(\mexpr,\menv),\mstore}\\
  \mktab \in \KTab &= (\Expr \times \Env)\times \Store \to \wp(\Kont) \\
  \mmemo \in \Memo &= (\Expr \times \Env)\times \Store \to \wp(\Value)
\end{align*}

\paragraph{The role of summaries:} notice the additional $\Memo$
component. A ``summary edge'' in CFA2 or equivalently an
$\epsilon$-edge in PDCFA is an edge from the source of a push edge to
the target of the matching pop edge. ``Matching'' here means there is
a path through machine reductions that don't change the stack, or
through summary edges. In our model, we only ``push'' when we call a
function, and we only ``pop'' when we return with a value. There is an
analogy to something more operational: summary edges embody
memoization. Instead of following an entire path through a call to
return a value, we simply jump from the call to the return (the result
of the call). We will return to this analogy and what it
means in the presence of first-class control in section \ref{sec:sr}.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\sclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
\\
    & $\tpl{\mval, \mstore', \mkont, \mktab, \mmemo} \text{ if } \mval \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

Once we finitize the address space of the store, this computes a
characteristic finite state machine that has only reachable states of
the pushdown system the unbounded stack model embodies.
To turn this semantics into PDCFA's (implemented) algorithm for constructing a Dyck
state graph, we apply a widening operator ${\mathcal F}$ to make $\mstore$, $\mkont$,
and $\mmemo$ shared amongst all states.

\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S', (\mastate',\mstore') \notin S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

A $\System$ embodies the states seen and at which store, $S$, the
frontier set of states yet to analyze, $F$, the shared store for the
frontier, $\mstore$, the continuation table $\mktab$ and the memo
table $\mmemo$. All the states in the frontier are stepped with the
current store, after which the next store is the least upper bound of
all the resulting stores. The next frontier contains only states
resulting from stepping the previous frontier that we haven't seen at
this next store. Systematic techniques for a performant implementation
can be found in \citet{ianjohnson:oaam:2013}.

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved nondeterminism}
\end{enumerate}
The first of these is an addition to the stackless context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have been
observed to decrease running time in most cases.

The second of these is to ameliorate a problem they call ``fake
rebinding.'' That is, since bindings in the abstract represent several
values, we don't want to reference a variable \texttt{x} in two
different places and have it resolve to two different values. In AAM,
a variable reference non-deterministically steps to all possible
values associated with that variable. Here we want to say that once
\texttt{x} is considered to stand for value $v$, then all subsequent
references of \texttt{x} should be $v$. If they aren't, it looks as if
\texttt{x} was rebound; it hasn't, and thus it is a ``fake
rebinding.'' CFA2 does not step to all values on variable reference,
but instead carries all its values around in superposition until they
need to be observed at, say, a function call. We give a simplified
semantics that is more along the AAM style, but CFA2's approach can
easily be recovered from it\footnote{Our Redex model implements the fake rebinding fix after function return}.

We show only the significantly modified rules of the semantics in
figure \ref{fig:frame-semantics}. The other rules simply carry along
the extra $\mframe$ component untouched.
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar[\mlab]{\mvar}, \menv), \mstore, \mframe, \mkont}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$ if $(\mframe', \mval) \in \lookup(\mstore,\mframe,\menv,\mvar,\mlab)$
    \\
% % application
%     $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mframe, \mkont}$
%     &
%     $\tpl{(\mexpri0, \menv), \mstore, \mframe, \kar{\mexpri1,\menv,\mkont}}$
%     \\
% % argument evaluation
%     $\tpl{\mval, \mstore, \mframe, \kar{\mexpr,\menv,\mkont}}$
%     &
%     $\tpl{(\mexpr, \menv), \mstore, \mframe, \kfn{\mval, \mkont}}$
%     \\
% function call
    $\tpl{\mval,\mstore,\mframe,\kfn{\sclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore', \mframe', \mkont}$
    \\ where & $(\mstore',\mframe') = \bind(\mstore,\mframe,\maddr,\mvar,\mval)$
  \end{tabular}
  \caption{The CES$\xi$K machine}
  \label{fig:frame-semantics}
\end{figure}

The metafunctions the semantics uses to entend and consult the heap
and stack use implicit information from the preanalysis we described
above:

@@

Sidestepping fake rebinding does not need to be restricted to
stackable references, but that is what CFA2 does. Indeed, as soon as
the nondeterminism has been determined, we could extend the stack
frame so any subsequent reference means what it meant previously in
the function. Lookup would then always try the stack frame before
falling back on the heap.

The widening we applied to PDCFA also applies here in order to get
CFA2. Both PDCFA and CFA2 use $\epsilon$-edges and summaries to
seemingly skip work when new stacks reach a function, but the summary
is not likely to be there, so we just drop that whole bit of
complexity from the semantics. CFA2 also included what they called
``transitive summaries'' to deal with tail calls. We insert an $\rt$
frame at every function call.  This view contends with tail calls, but
we can identify tail calls easily --- any call with an $\rt$ as its
continuation is a tail call. Tail calls are important in language
implementations for space complexity reasons
\citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these
concerns are less important. The repeated popping of $\rt$ inserted
by what otherwise were tail-calls is synonymous with CFA2's transitive
summaries.

\section{Analysis of delimited, composable control}
\label{sec:sr}

@@

The way we have been splitting continuations at function calls has
similarities to the meta-continuation approach to modeling delimited
control, given in figure \ref{fig:shift-reset} (adapted from
~\citep{ianjohnson:Biernacki2006274}). We could view each function call
as inserting a prompt, and returns as aborting to the nearest prompt.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% Reset
    $\tpl{(\sreset{\mexpr}, \menv), \mstore, \mkont, \mmkont}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kmt, \mkapp{\mkont}{\mmkont}}$
    \\
% Pop prompt
    $\tpl{\mval, \mstore, \kmt, \mkapp{\mkont}{\mmkont}}$
    &
    $\tpl{\mval, \mstore, {\mkont}, {\mmkont}}$
    \\
% Shift
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mkont},\kmt,\mmkont}$
    \\
% continuation call
    $\tpl{\mval,\mstore,\kfn{\mkont', \mkont}, \mmkont}$
    &
    $\tpl{\mval, \mstore, \mkont', \mkapp{\mkont}{\mmkont}}$
  \end{tabular}  
  \caption{Machine semantics for shift/reset}
  \label{fig:shift-reset}
\end{figure}
